---
title: "Exploration 1"
format: html
editor: visual
---

## Introduction

Objective: Visually explore how Menalopic Equivalent Daylight Illuminance (MEDI) and body movement (PIM) relate to each other.

## Prepare data

### Load dependencies

```{r}
library(tidyverse)
library(melidosData)
library(zoo)
library(gridExtra)
library(lubridate)
```

### (Down)load data

Load Melidos Data from all sites and store locally.

```{r}
modality = "light_wrist"

outputfile = paste0(modality, ".RData")
if (!file.exists(outputfile)) {
  # Only download data when not downloaded before
  data <- melidosData::load_data(modality = modality)
  save(data, file = outputfile)
} else {
  load(file = outputfile)
}
```

### Inspect sample sizes

```{r}
Nall = 0
for (i in 1:(length(data))) {
  N = length(unique(data[[i]]$Id))
  cat(paste0("N ", names(data)[[i]], ": ", N, "\n"))
  Nall = Nall + N
}
cat(paste0("N all ", Nall, "\n"))
```

### Collapse all sites into a single tibble

```{r}
D = purrr::map(data, bind_rows, .id = "name") |> 
  dplyr::bind_rows(.id = "site") |>
  dplyr::select(Id, Datetime, MEDI, PIM, site, name)
dplyr::glimpse(D)
```

### Keep only complete cases

```{r}
D = D |> ungroup() |>
  dplyr::filter(if_all(c(Datetime, MEDI, PIM), ~!is.na(.x)))

dplyr::glimpse(D)
```

### Derivative for PIM and MEDI

```{r}
D = D |>
  mutate(delta_PIM = c(0, diff(PIM))) |>
  mutate(delta_MEDI = c(0, diff(MEDI)))
```

## Relation MEDI versus PIM

Note that data includes all time series from all individuals in all study sites. Therefore, data is shown as a density plot.

```{r}
    ggplot(D, aes(x = MEDI, y = PIM)) +
      geom_bin2d(binwidth = c(4000, 1000), aes(fill = after_stat(count)),
                 color = "black", alpha = 0.8) +
      scale_fill_gradient(name = "Density", low = "lightgreen", high = "darkgreen") +
      labs(title = "Density of MEDI vs PIM", x = "MEDI", y = "PIM") +
      facet_wrap(~site) +
      theme_minimal() +
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL)
```

## Density of change in MEDI versus change in PIM

Another way of looking at the data is by comparing change in movement and light exposure.

```{r}
  myplot = function(D) {  
    # Define  custom plot function to show relation between variables
    ggplot(D, aes(x = delta_MEDI, y = delta_PIM)) +
      geom_bin2d(binwidth = c(4000, 1000), aes(fill = after_stat(count)),
                 color = "black", alpha = 0.8) +
      scale_fill_gradient(name = "Density", low = "lightblue", high = "darkblue") +
      labs(title = "Density of change in MEDI vs change in PIM",
           x = "Change in in MEDI", y = "Change in PIM") +
      facet_wrap(~site) +
      theme_minimal() +
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL)
  }
  myplot(D)
```

## Inspect specific conditions

It may be worth repeating this plost for specific conditions.

### Only time segments with high correlation between PIM and MEDI?

Derive correlation between PIM and MEDI for each 5 minute window

```{r}
n_epoch_per_minute = 6
test = zoo::rollapply(D[, c("MEDI", "PIM")],
                 width = n_epoch_per_minute * 5, by = n_epoch_per_minute * 5, function(x) ifelse(sd(x[,1]) != 0 && sd(x[,2]) != 0, cor(x[, 1], x[, 2], use = "complete.obs"), 0), by.column = FALSE)
D$corr = c(rep(test, each = n_epoch_per_minute * 5), rep(0, 15))

myplot(D |> filter(corr > 0.6))
```

### Only daytime (9:00 - 20:00)?

What if association is only present during the day?

```{r}
D <- D |> mutate(hour = lubridate::hour(Datetime) + lubridate::minute(Datetime)/60)
myplot(D |> filter(hour >= 9 & hour <= 20))
```

### What if we only look at night (23:00 - 7:00)?

What if association is only present during the night?

```{r}
D <- D |> mutate(hour = lubridate::hour(Datetime) + lubridate::minute(Datetime)/60)
myplot(D |> filter(hour <= 7 | hour >= 23))
```

### Only summer data (June-August)?

What if association is only present during the summer?

```{r}
D <- D |> mutate(month = lubridate::month(Datetime))
myplot(D |> filter(month >= 6 & month <= 8))
```

### Only winter data (December-February)?

What if association is only present during the winter?

```{r}
D <- D |> mutate(month = lubridate::month(Datetime))
myplot(D |> filter(month <= 2 | month >= 12))
```

## What if data transformations are needed?

### Simplify MEDI to categorical values?

```{r}
D <- D |> dplyr::mutate(MEDI_LEVEL = cut(MEDI,
                                  breaks = c(0, 1, 10, 100, 1000, 10000, 100000, Inf)))

D <- D |> mutate(delta_MEDI_LEVEL = c(0, diff(MEDI_LEVEL)))

ggplot(D, aes(x = delta_MEDI_LEVEL, y = delta_PIM)) +
      geom_bin2d(binwidth = c(1, 1000), aes(fill = after_stat(count)),
                 color = "black", alpha = 0.8) +
      scale_fill_gradient(name = "Density", low = "lightblue", high = "darkblue") +
      labs(title = "Density of change in MEDI level vs change in PIM",
           x = "Change in in MEDI level ", y = "Change in PIM") +
      facet_wrap(~site) +
      theme_minimal() +
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL)
```

### Simplify PIM to categorical values?

```{r}
D <- D |> dplyr::mutate(PIM_LEVEL = cut(PIM,
                                  breaks = c(0, 10, 100, 1000, 10000, 100000, Inf)))

D <- D |> mutate(delta_PIM_LEVEL = c(0, diff(PIM_LEVEL)))

ggplot(D, aes(x = delta_MEDI, y = delta_PIM_LEVEL)) +
      geom_bin2d(binwidth = c(4000, 1), aes(fill = after_stat(count)),
                 color = "black", alpha = 0.8) +
      scale_fill_gradient(name = "Density", low = "lightblue", high = "darkblue") +
      labs(title = "Density of change in MEDI vs change in PIM level",
           x = "Change in in MEDI", y = "Change in PIM level") +
      facet_wrap(~site) +
      theme_minimal() +
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL)
```

### Downsample PIM and MEDI to 1 minute resolution?

```{r}
D$rounded_time_minutes = round(as.numeric(D$Datetime) / 60) * 60

myagg = function(x) {
  if (inherits(x, "numeric")) {
    return(mean(x))
  } else {
    return(x[1])
  }
}
Dtmp <- as_tibble(aggregate(D, by = list(D$rounded_time_minutes, D$Id), FUN = myagg)) |>
  select(-c(Group.1, Group.2))

```

```{r}
ggplot(Dtmp, aes(x = delta_MEDI, y = delta_PIM)) +
      geom_bin2d(binwidth = c(2000, 1000), aes(fill = after_stat(count)),
                 color = "black", alpha = 0.8) +
      scale_fill_gradient(name = "Density", low = "lightblue", high = "darkblue") +
      labs(title = "Density of change in MEDI vs change in PIM level",
           x = "Change in in MEDI", y = "Change in PIM level") +
      facet_wrap(~site) +
      theme_minimal() +
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL)
```


## Reflections

- No clear associations between PIM and MEDI or change in PIM and change in MEDI.
- No clear association when focussing only on day or night, summer or winter.
- No clear association after changing the resolution in values or in time.
- Possible reasons:
  - All data per site combined, zooming in on a smaller subset or even individual recordings may provide more insight.
  - Only looked at isolated effect of each factor, not at the combined effect.
  - Limited attempt to clean data
  - Only focused on epoch level comparisons





