---
title: "Exploration 2b"
format: html
editor: visual
---

## Introduction

Same as exploration 2 but now with MEDI from eyel-level and PIM from wrist,
and in UCR instead of TUM as for TUM there is only one Id for which we have matching
MEDI glasses and PIM wrist, while UCR has 6.

## Prepare data

### Load dependencies

```{r}
library(tidyverse)
library(melidosData)
library(gridExtra)
library(lubridate)
```

### (Down)load data

Load Melidos Data from all sites and store locally.

```{r}
for (modality in c("light_wrist", "light_glasses")) {
  site = "UCR"
  outputfile = paste0(modality, "_", site, ".RData")
  if (!file.exists(outputfile)) {
    # Only download data when not downloaded before
    data <- melidosData::load_data(modality = modality, site = site)
    save(data, file = outputfile)
  } else {
    load(file = outputfile)
  }
  if (modality == "light_wrist") {
    data_light_wrist = data
  } else {
    data_light_glasses = data
  }
}

# combine into one data object with MEDI from glasses and PIM from wrist 
data = merge(data_light_glasses,
                           data_light_wrist,
                           by = c("Datetime", "Id"),
                           suffixes = c(".glass", ".wrist"))
data <- data |>
  dplyr::select(Id, Datetime, MEDI.glass, PIM.wrist) |>
  rename(MEDI = MEDI.glass, PIM = PIM.wrist)
```

### Inspect sample sizes

```{r}
Nall = 0
N = length(unique(data$Id))
cat(paste0("N ", site, ": ", N, "\n"))
```

### Select only relevant columns

```{r}
D = data |> dplyr::select(Id, Datetime, MEDI, PIM)
dplyr::glimpse(D)
```

### Keep only complete cases

```{r}
D = D |> ungroup() |>
  dplyr::filter(if_all(c(Datetime, MEDI, PIM), ~!is.na(.x)))

dplyr::glimpse(D)
```

## Average 24 hours cycles

Derive average 24 cycle by aggregating values for identical clock times across days.

```{r}
D <- D |> mutate(clocktime = paste(lubridate::hour(Datetime),
                                   lubridate::minute(Datetime),
                                   lubridate::second(Datetime), sep = ":"),
                 hour_in_day = lubridate::hour(Datetime) +
                   (lubridate::minute(Datetime) / 60) +
                   (lubridate::second(Datetime) / 3600)) |>
  arrange(Id, hour_in_day)

myagg = function(x) {
  if (inherits(x, "numeric")) {
    return(mean(x))
  } else {
    return(x[1])
  }
}
D24 <- as_tibble(aggregate(D, by = list(D$clocktime, D$Id), FUN = myagg))
D24 = D24 |> select(-c(Group.1, Group.2, Datetime, clocktime)) |> arrange(Id, hour_in_day)
```

```{r}
scaleFactor <- max(D24$PIM) / max(D24$MEDI)
ggplot(D24, aes(x=hour_in_day)) +
  geom_line(aes(y=PIM), col="blue") +
  geom_line(aes(y=MEDI * scaleFactor), col="red") +
  scale_y_continuous(name="PIM", sec.axis=sec_axis(~./scaleFactor, name="MEDI")) +
  theme(
    axis.title.y.left=element_text(color="blue"),
    axis.text.y.left=element_text(color="blue"),
    axis.title.y.right=element_text(color="red"),
    axis.text.y.right=element_text(color="red")
  ) +
  facet_wrap(~Id)
```
### With smoothing?

```{r}
scaleFactor <- max(D24$PIM) / max(D24$MEDI)
ggplot(D24, aes(x=hour_in_day)) +
  geom_smooth(aes(y=PIM), method="gam", col="blue") +
  geom_smooth(aes(y=MEDI * scaleFactor), method="gam", col="red") +
  scale_y_continuous(name="PIM", sec.axis=sec_axis(~./scaleFactor, name="MEDI")) +
  theme(
    axis.title.y.left=element_text(color="blue"),
    axis.text.y.left=element_text(color="blue"),
    axis.title.y.right=element_text(color="red"),
    axis.text.y.right=element_text(color="red")
  ) +
  facet_wrap(~Id)
```


## Explore change points - low PIM => high PIM

In this section I am putting most code in functions to ease repeating the
same analysis for high PIM => low PIM, low MEDI => high MEDI, high MEDI => low MEDI.

### Go back to original data object

```{r}
rm(D)
D = data |> dplyr::select(Id, Datetime, MEDI, PIM) |>
  ungroup() |>
  dplyr::filter(if_all(c(Datetime, MEDI, PIM), ~!is.na(.x))) |>
  arrange(Id, Datetime)
D_backup = D # Keep backup for next section
```

### Find time points where value transitions

I am using an improvised change point detection algorithm.

```{r}
PIM_change_detection = function(x){
  # x = sequence
  change = FALSE
  N = length(x)
  # ignore incomplete time series at start and end
  if (N < 60) return(change)
  midpoint = ceiling(N/2)
  mn_before = mean(x[1:midpoint])
  sd_before = sd(x[1:midpoint])
  max_before = max(x[1:midpoint])
  mn_after = mean(x[(midpoint + 1):N])
  # heuristic logic to identify changes from inactive to active
  if (mn_after > mn_before * 10 && 
      mn_before < 200 &&
      sd_before < 100 &&
      max_before < 100 &&
      mn_after > 500) { 
    # use t-test to test for actual difference (note: maybe redundant)
    tt = t.test(x[1:midpoint], x[(midpoint + 1):N])
    if (tt$p.value < 0.01) { # t-test significant
      change = TRUE
    }
  }
  return(change) # boolean: true = change, false = no change
}
epoch_size = 10
half_window_size = (60 / epoch_size) * 5 # 5 minutes
D$change_points = 0
for (id in unique(D$Id)) {
  id_segment = which(D$Id == id)
  D$change_points[id_segment] = slider::slide_dbl(.x = D$PIM[id_segment],
                                      .f = ~PIM_change_detection(.x),
                                      .before = half_window_size,
                                      .after = half_window_size,
                                      .complete = FALSE)
}
```

### Find middle of each sequence labelled as changepoint

The above results in sequences of change points, for convenience select middle point of such sequence only.

```{r}
find_middle = function(D) {
  D$change_points_final = 0
  for (id in unique(D$Id)) {
    S = D[which(D$Id == id), ]
    K = rle(S$change_points)
    pos = 0
    change_point_positions = c()
    for (i in 1:length(K$values)) {
      if (K$values[i] == 0) {
        pos = pos + K$lengths[i]
      } else {
        pos = pos + ceiling(K$lengths[i] / 2) # add half to reach midpoint
        change_point_positions = c(change_point_positions, pos)
        pos = pos + floor(K$lengths[i] / 2) # add other half
      }
    }
    S$change_points_final[change_point_positions] = 1
    D$change_points_final[which(D$Id == id)] = S$change_points_final
  }
  return(D)
}
D = find_middle(D)
```

### Create tibble with all time series centered around the change points

To be able to plot the data we need a long format object with all time series before and after the change points.

```{r}
create_tibble = function(D, half_window_size, epochsize) {
  output = c()
  for (id in unique(D$Id)) {
    S = D[which(D$Id == id),]
    N = nrow(S)
    change_points = which(S$change_points_final == 1)
    # initialise tibble to store all the sequences
    Nc =  length(change_points) * ((half_window_size * 2) + 1)
    ts = tibble(changepoint = numeric(Nc),
                MEDI = numeric(Nc),
                PIM = numeric(Nc),
                time = numeric(Nc),
                Id = character(Nc))
    ts$Id = id
    i = 1
    if (length(change_points) > 0) {
      for (j in change_points) {
        if (j > half_window_size + 1 && (j + half_window_size) <= N) {
          segment = i:(i + (half_window_size * 2))
          ts$PIM[segment] = S$PIM[(j - half_window_size):(j + half_window_size)]
          ts$MEDI[segment] = S$MEDI[(j - half_window_size):(j + half_window_size)]
          ts$time[segment] = -half_window_size:half_window_size / (60/epoch_size)
          ts$changepoint[segment] = j
          i = i + 1 + (half_window_size * 2)
        }
      }
      output = rbind(output, ts)
    }
  }
  output <- output |> group_by(changepoint)
  return(output)
}
output = create_tibble(D, half_window_size, epochsize)
```

### Plot

```{r}
plot_changepoint_ts = function(output, title = "") {
  scaleFactor <- max(output$PIM) / max(output$MEDI)
  ggplot(output, aes(x = time, group = changepoint)) +
    geom_line(aes(y = PIM), col = "blue") +
    geom_line(aes(y = MEDI * scaleFactor), col = "red") +
    scale_y_continuous(name = "PIM", sec.axis = sec_axis(~./scaleFactor, name = "MEDI")) +
    theme(
      axis.title.y.left = element_text(color = "blue"),
      axis.text.y.left = element_text(color = "blue"),
      axis.title.y.right = element_text(color = "red"),
      axis.text.y.right = element_text(color = "red")
    ) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "time (minutes)") + 
    facet_wrap(~Id) + ggtitle(title)
  
}
plot_changepoint_ts(output, "low PIM => high PIM")
```

## Explore change points - high PIM => low PIM

Now to the same as above, but from a MEDI perspective

### Go back to original data object

```{r}
D = D_backup
```

### Find time points where value transitions

I am using an improvised change point detection algorithm.

```{r}
PIM_change_detection = function(x){
  # x = sequence
  change = FALSE
  N = length(x)
  # ignore incomplete time series at start and end
  if (N < 60) return(change)
  midpoint = ceiling(N/2)
  mn_before = mean(x[1:midpoint])
  sd_after = sd(x[(midpoint + 1):N])
  max_after = max(x[(midpoint + 1):N])
  mn_after = mean(x[(midpoint + 1):N])
  # heuristic logic to identify changes from inactive to active
  if (mn_before > mn_after * 10 && 
      mn_after < 2000 &&
      sd_after < 100 &&
      max_after < 100 &&
      mn_before > 500) { 
    # use t-test to test for actual difference (note: maybe redundant)
    tt = t.test(x[1:midpoint], x[(midpoint + 1):N])
    if (tt$p.value < 0.01) { # t-test significant
      change = TRUE
    }
  }
  return(change) # boolean: true = change, false = no change
}
epoch_size = 10
half_window_size = (60 / epoch_size) * 5 # 5 minutes
D$change_points = 0
for (id in unique(D$Id)) {
  id_segment = which(D$Id == id)
  D$change_points[id_segment] = slider::slide_dbl(.x = D$PIM[id_segment],
                                      .f = ~PIM_change_detection(.x),
                                      .before = half_window_size,
                                      .after = half_window_size,
                                      .complete = FALSE)
}
```

However, this results in sequences of change points as multiple time points may meet the change point criteria. For convenience select middle point of such sequence.

### Repeat same processing steps as before

```{r}
D = find_middle(D)
output = create_tibble(D, half_window_size, epochsize)
```

### Plot

```{r}
plot_changepoint_ts(output, "high PIM => low PIM")
```


## Explore change points - low MEDI => high MEDI

Now to the same as above, but from a MEDI perspective

### Go back to original data object

```{r}
D = D_backup
```

### Find time points where value transitions

I am using an improvised change point detection algorithm.

```{r}
MEDI_change_detection = function(x){
  # x = sequence
  change = FALSE
  N = length(x)
  # ignore incomplete time series at start and end
  if (N < 60) return(change)
  midpoint = ceiling(N/2)
  mn_before = mean(x[1:midpoint])
  sd_before = sd(x[1:midpoint])
  max_before = max(x[1:midpoint])
  mn_after = mean(x[(midpoint + 1):N])
  # heuristic logic to identify changes from inactive to active
  if (mn_after > mn_before * 10 && 
      mn_before < 1000 &&
      sd_before < 200 &&
      max_before < 200 &&
      mn_after > 1000) { 
    # use t-test to test for actual difference (note: maybe redundant)
    tt = t.test(x[1:midpoint], x[(midpoint + 1):N])
    if (tt$p.value < 0.01) { # t-test significant
      change = TRUE
    }
  }
  return(change) # boolean: true = change, false = no change
}
epoch_size = 10
half_window_size = (60 / epoch_size) * 5 # 5 minutes
D$change_points = 0
for (id in unique(D$Id)) {
  id_segment = which(D$Id == id)
  D$change_points[id_segment] = slider::slide_dbl(.x = D$MEDI[id_segment],
                                      .f = ~MEDI_change_detection(.x),
                                      .before = half_window_size,
                                      .after = half_window_size,
                                      .complete = FALSE)
}
```

However, this results in sequences of change points as multiple time points may meet the change point criteria. For convenience select middle point of such sequence.

### Repeat same processing steps as before

```{r}
D = find_middle(D)
output = create_tibble(D, half_window_size, epochsize)
```

### Plot

```{r}
plot_changepoint_ts(output, "low MEDI => high MEDI")
```

## Explore change points - high MEDI => low MEDI

Now to the same as above, but from a MEDI perspective

### Go back to original data object

```{r}
D = D_backup
```

### Find time points where value transitions

I am using an improvised change point detection algorithm.

```{r}
MEDI_change_detection = function(x){
  # x = sequence
  change = FALSE
  N = length(x)
  # ignore incomplete time series at start and end
  if (N < 60) return(change)
  midpoint = ceiling(N/2)
  mn_before = mean(x[1:midpoint])
  sd_after = sd(x[(midpoint + 1):N])
  max_after = max(x[(midpoint + 1):N])
  mn_after = mean(x[(midpoint + 1):N])
  # heuristic logic to identify changes from inactive to active
  if (mn_before > mn_after * 10 && 
      mn_after < 1000 &&
      sd_after < 200 &&
      max_after < 200 &&
      mn_before > 1000) { 
    # use t-test to test for actual difference (note: maybe redundant)
    tt = t.test(x[1:midpoint], x[(midpoint + 1):N])
    if (tt$p.value < 0.01) { # t-test significant
      change = TRUE
    }
  }
  return(change) # boolean: true = change, false = no change
}
epoch_size = 10
half_window_size = (60 / epoch_size) * 5 # 5 minutes
D$change_points = 0
for (id in unique(D$Id)) {
  id_segment = which(D$Id == id)
  D$change_points[id_segment] = slider::slide_dbl(.x = D$MEDI[id_segment],
                                      .f = ~MEDI_change_detection(.x),
                                      .before = half_window_size,
                                      .after = half_window_size,
                                      .complete = FALSE)
}
```

### Repeat same processing steps as before

```{r}
D = find_middle(D)
output = create_tibble(D, half_window_size, epochsize)
```

### Plot

```{r}
plot_changepoint_ts(output, "high MEDI => low MEDI")
```


## Reflections

### 24 hour averaged data

Same as for exploration 2:

- Non-smoothed 24 hour averaged MEDI and PIM shows that light exposure and activity tend to happen during the same hours of the day, but do not reveal clear general patterns.
- Smoothed 24 hour averaged MEDI and PIM shows that they sometimes be correlated and sometimes not, with large variation between individuals.


### Change points

Partially the same as for exploration 2 where we observate that transtions from low -> high or high -> low are associated. However, for some visualisation it looks as if the relation is now reversed: PIM going down being associated with MEDI going up. Better visualisations are needed to gain more insight.

